---
title: "ERBOT: A Unified Entity Resolution Pipeline"
subtitle: "Vignette — end-to-end introduction, design, and worked example"
author: "Xinmin Chu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ERBOT: A Unified Entity Resolution Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## 1. Introduction

**Entity resolution** (ER) — also called deduplication, record linkage, or
entity matching — answers a deceptively simple question: *which records in my
data refer to the same real-world entity?*  Customer databases contain "Acme
Corp" and "ACME Corporation"; bibliographic databases contain the same paper
cited with slightly different author lists; product catalogs contain the same
item under different SKUs.  ERBOT is an R package that solves this problem
end-to-end through a **nine-stage pipeline**, from raw table to cluster labels,
with honest evaluation built in.

### Design principles

1. **"Data in, entity out."**  One call — `er_run()` — takes any tabular
   dataset and returns cluster labels.  Every stage is also usable
   standalone for researchers who want to swap components.
2. **NA-aware everything.**  Real data have missing values.  Similarities,
   weights, and combination all handle `NA` explicitly instead of silently
   dropping or imputing it.
3. **Honest evaluation by default.**  Ground truth, when supplied, is used
   *only* for evaluation — never for model selection (choice of *k*, choice
   of clustering method) — unless the user explicitly opts into
   `supervised_selection = TRUE`.  This discipline is enforced in code, not
   just documented.
4. **Method coverage over method advocacy.**  Eight unsupervised clustering
   methods plus 11 supervised classifiers, five
   weight-learning strategies, four blocking methods, and an ablation
   framework that compares them fairly.  The package helps you find *which*
   method works for *your* data, not which method the authors prefer.

## 2. Installation

```{r install, eval = FALSE}
# Core dependencies
install.packages(c(
  "data.table", "dplyr", "tibble", "stringr", "stringi", "tidyr",
  "purrr", "readr", "readxl", "text2vec", "Matrix", "irlba",
  "stringdist", "igraph", "FNN", "cluster", "scales",
  "grid", "gridExtra", "magrittr", "jsonlite"
))

# Optional but recommended
install.packages("aricode")                  # ARI and related metrics
remotes::install_github("ddegras/GCMER")     # agreement metrics & graph coloring

# Install erbot
remotes::install_github("xinminchu/erbot", upgrade = "never")
```

R ≥ 4.1 is required.

## 3. Quick start

```{r quickstart, eval = FALSE}
library(erbot)

# Any data.frame, file path, or benchmark keyword ("cora", "affiliation", "d10k")
res <- er_run(
  data   = my_data,
  truth  = my_truth,      # optional: named integer vector, or data.frame(id, cluster_id)
  verbose = TRUE
)

res$clusters      # integer vector of final cluster labels (length n)
res$performance   # tibble of ARI, B-cubed F, homogeneity, completeness, ...
res$predictions   # tibble(id, cluster_id), ready for write-out
print(res)        # compact summary
```

With `out_dir` set, `er_run()` also writes `er_predictions.csv` and
`er_performance.csv` to disk.

## 4. The nine-stage pipeline

Every `er_run()` call executes the same nine stages.  Understanding them is
understanding the package.

### Stage 1 — Load (`er_load()`)

Ingests the input, which may be a `data.frame`, a file path (CSV/TSV/Excel),
or a benchmark keyword (`"cora"`, `"affiliation"`, `"d10k"`).  Column names
are normalized to lowercase; the result is a tibble ready for diagnosis.

### Stage 2 — Diagnose (`er_diagnose()`)

Profiles every field and decides *how* the pipeline should treat it:

- **ID column** auto-detection (candidates: `id`, `record_id`, `paper_id`, …).
- **Mode** detection: `dedup` (one table) vs `link` (two sources, detected
  via a `source_id` column).
- **Field typing**: for each column, the missing-value rate, cardinality, and
  a recommended similarity method (`jw` for short strings, `cosine` for long
  text, `numeric` for numbers, `exact` for codes).
- **Blocking recommendation**: whether blocking is needed at all
  (`blocking_needed`), and which method/key to use.

The diagnosis object drives all downstream `"auto"` defaults, so a user who
changes nothing still gets a sensible configuration.

### Stage 3 — Block (`er_block()`)

Naive pairwise comparison is O(n²).  Blocking generates a *candidate pair*
set that is orders of magnitude smaller while (hopefully) retaining all true
matches.  Four methods:

| Method | Idea |
|---|---|
| `"none"` | All pairs (only for small n; guarded by `max_pairs`) |
| `"standard"` | Exact match on a blocking key |
| `"prefix"` | Shared prefix of length `prefix_len` (default 3) |
| `"sn"` | Sorted neighbourhood with window `sn_window` (default 40) |

`er_block_stats()` reports the **reduction ratio** (fraction of pairs pruned)
and, when truth is available, **pair completeness** (fraction of true matches
retained).  A block that prunes 99% of pairs but loses 30% of true matches is
a bad block — this function makes that visible.

A built-in guard estimates C(m, 2) before allocating: if the candidate count
would exceed `max_pairs` (default 2M), blocking refuses to produce an
unbounded pair set instead of crashing with OOM.

### Stage 4 — Similarity (`er_similarity()`)

Computes per-field similarities **only on candidate pairs**, stored as a
named list of numeric vectors (`sim_list`).  Key properties:

- **NA-aware**: if either value is missing, the field contributes no signal
  rather than a spurious 0.  (The B-01 class of bugs — NA/empty-string
  crashes in similarity code — was fixed in the 2026-10-06 forging pass.)
- **Per-field methods**: Jaro–Winkler (`jw`) for names, TF-IDF cosine for
  long text, normalized numeric distance, exact match for codes.  The spec
  can come from `er_diagnose()` or be supplied by hand.

### Stage 5 — Weights (`er_weights()`)

Not all fields are equally informative.  Five strategies:

| Method | Weight ∝ | Needs truth? |
|---|---|---|
| `"equal"` | 1/K | No |
| `"fellegi_sunter"` | log(m_k / u_k) via EM | No |
| `"ari"` | per-field ARI vs truth | **Yes** |
| `"bimodal"` | bimodality of the similarity distribution | No |
| `"variance"` | variance of the similarity distribution | No |

`"auto"` picks `"ari"` when truth is supplied, `"fellegi_sunter"`
otherwise.  The ARI method supports nested inner CV (`inner_k`) so that
using ARI both for weight learning and for final evaluation does not leak.

### Stage 6 — Combine (`er_combine()`)

NA-aware weighted combination of the per-field similarities into a single
score per pair, then `er_pairs_to_sparse()` builds the symmetric sparse
similarity matrix **S** (n × n, `dgCMatrix`).  Everything downstream works
from **S**.

### Stage 7 — Cluster (`er_cluster_all()`)

Eight unsupervised clustering methods run on **S**, plus 11 supervised
classifiers when truth is available:

- **Centroidal** (need *k*): `hclust_avg`, `hclust_ward`, `pam`.  Features
  are derived from **S** via SVD (exact dense SVD for small n, `irlba`
  otherwise); *k* is auto-tuned over `k_grid` by silhouette when not given.
- **Graph-based**: `louvain`, `leiden`, `threshold_cc` (connected components
  on pairs above threshold), `label_prop` (label propagation), `gc` (graph
  coloring).
- **Supervised** (need truth): the 11 classifiers from `er_supervised_classifiers()`
  — logistic (Firth penalized by default, so separation doesn't diverge),
  LDA, QDA, KNN, weighted KNN, tree, random forest, xgboost, nnet,
  Fellegi–Sunter, radial SVM — trained on pair features
  and used to classify pairs before clustering.

A sparsity guard (`.check_dense_n()`, default n = 5000) prevents accidental
dense-matrix blowups in HC/PAM/GC on large data.

### Stage 8 — Merge (`er_merge()`)

Post-processing of the clustering ensemble:

- `"consensus"` (default): keeps pairs on which a fraction
  (`consensus_alpha`) of methods agree — unsupervised.
- `"transitivity"`: connected components on pairs with S ≥ threshold.
- `"best"`: picks the method with highest ARI vs truth — **supervised**,
  warns loudly, and only sees truth under `supervised_selection = TRUE`.
- `"none"`: skip.

### Stage 9 — Evaluate (`er_evaluate()`)

When truth is supplied, a 14-column metric table per method:

- **External**: ARI, AMI, B-cubed precision/recall/F, pairwise
  precision/recall/F, homogeneity, completeness, V-measure.
- Label alignment is done **by record name**, never by position — the
  silent-misalignment bug (B-04, fixed 2026-10-06) that once invalidated all
  reported numbers cannot recur: a length check against the ID vector is
  enforced.

Without truth, `er_unsupervised_quality()` reports within/between
similarity, the similarity gap, and silhouette.

### Stage 10 — Output

`er_run()` returns an `er_result` object (with a `print` method) containing
clusters, all per-method clusters, the performance table, the diagnosis,
blocking stats, weights, and parameters.  With `out_dir`, predictions and
performance are written to CSV.

## 5. Honesty discipline: how the package prevents fooling yourself

This is the part the 2026-10-06 forging pass was about, and it deserves its
own section because most ER toolkits get it wrong.

1. **Truth never drives model selection by default.**
   `supervised_selection = FALSE` is the default.  Truth is used for
   *evaluation* (stage 9) and optionally for *weight learning*
   (`weights = "ari"`), but never for choosing *k* or picking the best
   method.  `merge = "best"` without the flag warns and is blocked from
   seeing truth.
2. **Entity-level splits, not record-level splits.**  `er_split()`,
   `er_cv()`, and `er_protocol_b()` split by *entity* (all records of an
   entity go to the same fold).  Record-level splits leak near-duplicates
   across train/test and inflate scores.
3. **True held-out evaluation.**  `er_protocol_b()` partitions entities into
   Selection and Test sets; *k* and method are chosen on Selection
   (with inner entity-level CV); the Test set is clustered and evaluated
   exactly once.
4. **Seeded reproducibility.**  All stochastic steps take seeds; RNG state
   is saved and restored (`.rng_save()`) so a pipeline run doesn't perturb
   the caller's random stream.

## 6. Supporting frameworks

Beyond the pipeline, the package ships research infrastructure:

- **Ablation** (`er_ablation_all()`, `er_ablation_table()`): full factorial
  over weight conditions (uniform / Fellegi–Sunter EM / full-data ARI /
  CV-learned ARI) × blocking conditions (none / blocked), with
  `print_ablation()` summaries.
- **Tuning** (`er_tune()`): honest three-way train/validation/test protocol
  for hyperparameter search; internal vs external objectives.
- **CV protocols** (`er_protocol_a()`, `er_protocol_b()`): the "honest"
  evaluation protocols used in the paper.
- **Supervised pair classification** (`er_pair_features()`,
  `er_pair_classify()`, `er_supervised_classifiers()`): 11 classifiers with
  a validity gate — structurally unsuitable configurations return
  `valid = FALSE` with a reason instead of silently producing garbage.
- **Baselines** (`er_baseline_splink()`, `er_baseline_deepmatcher()`,
  `er_baseline_ditto()`): wrappers that call out to Python for
  probabilistic (Splink) and deep-learning (DeepMatcher, Ditto) baselines.
- **Scenarios** (`er_list_scenarios()`, `er_run_scenario()`): 21 preset
  configurations with friendly error messages, built for teaching and
  demos.
- **Reporting** (`er_save_report_pdf()`, `er_draw_table()`): PDF reports
  with performance tables and diagnostic plots.

## 7. Worked example

A tiny self-contained deduplication task: 12 records, 4 true entities
(each with small name variations).

```{r worked-example, eval = FALSE}
library(erbot)

toy <- data.frame(
  id   = paste0("r", 1:12),
  name = c("Acme Corporation", "Acme Corp", "ACME Corp.",
           "Beacon Technologies", "Beacon Tech", "Beacon Technology Inc",
           "Crestline Labs", "Crestline Lab", "Crestline Laboratories",
           "Driftwood Systems", "Driftwood Sys", "Driftwood System Inc"),
  stringsAsFactors = FALSE
)
truth <- setNames(rep(1:4, each = 3), toy$id)

res <- er_run(
  data            = toy,
  truth           = truth,
  cluster_methods = c("louvain", "threshold_cc", "hclust_avg"),
  verbose         = FALSE
)

print(res)
#> er_result: n=12 records, 4 final clusters
#> Performance:
#>   method         ARI Bcubed_F Vmeasure
#>   louvain      1.000    1.000    1.000
#>   threshold_cc 0.000    0.400    0.000
#>   hclust_avg   0.891    0.941    0.946
#>   final        1.000    1.000    1.000

# Inspect what the pipeline decided at each stage
res$diag$recommended_block_method   # which blocking was chosen
res$weights                         # learned field weights
res$blocking$stats$reduction_ratio   # fraction of pairs pruned
res$predictions                     # id -> cluster_id
```

Note how the three methods disagree (`threshold_cc` collapses here at the
default threshold, `hclust_avg` is close but not perfect) yet the consensus
merge (`final`) recovers the correct 4 clusters — this is exactly why the
package runs an ensemble instead of betting on one method.

Stage-by-stage manual control is equally possible:

```{r manual-stages, eval = FALSE}
df    <- er_load(toy)
diag  <- er_diagnose(df)
pairs <- er_block(df, diag = diag, method = "prefix", prefix_len = 3)
sim   <- er_similarity(df, pairs, diag = diag)
wt    <- er_weights(sim, method = "fellegi_sunter")
S     <- er_pairs_to_sparse(pairs, er_combine(sim, weights = wt), nrow(df))
labs  <- er_cluster(S, method = "louvain")
er_evaluate(list(louvain = labs), truth = truth, id_vec = df$id)
```

## 8. Function reference map

| Task | Entry points |
|---|---|
| Full pipeline | `er_run()` |
| Load / diagnose | `er_load()`, `er_diagnose()`, `er_data_profile()` |
| Blocking | `er_block()`, `er_block_stats()` |
| Similarity | `er_similarity()`, `er_pairwise_stringdist()`, `er_tfidf_svd()` |
| Weights | `er_weights()`, `er_combine()`, `er_pairs_to_sparse()` |
| Clustering | `er_cluster()`, `er_cluster_all()`, `er_merge()`, `er_consensus()`, `er_transitivity()` |
| Evaluation | `er_evaluate()`, `er_bcubed()`, `er_vmeasure()`, `er_unsupervised_quality()` |
| Honest protocols | `er_protocol_a()`, `er_protocol_b()`, `er_cv()`, `er_split()`, `er_tune()` |
| Ablation | `er_ablation_all()`, `er_ablation_table()`, `print_ablation()` |
| Supervised | `er_pair_features()`, `er_pair_classify()`, `er_supervised_classifiers()` |
| Baselines | `er_baseline_splink()`, `er_baseline_deepmatcher()`, `er_baseline_ditto()` |
| Scenarios/demos | `er_list_scenarios()`, `er_run_scenario()`, `er_scenario_presets()` |
| Reporting | `er_save_report_pdf()`, `er_draw_table()`, `print.er_result()` |
| Datasets | `er_build_restaurant()`, `er_build_dblp_acm()`, `er_load_gold()` |

## 9. Notes for the 2026-10-09 advisor meeting

What changed since the 10-02 meeting (the forging pass, all verified —
199 tests, 0 failures, 0 warnings, run by Sam on Windows):

- **Phase 1 (blockers)**: NA/empty-string crashes fixed; evaluation label
  alignment fixed (old numbers were silently wrong); blocking OOM guard;
  ARI-on-disjoint-sets rewritten; DESCRIPTION/NAMESPACE/LICENSE consistent.
- **Phase 2 (honesty)**: `supervised_selection` flag; truth no longer drives
  *k*-tuning or method choice by default; `er_tune()` and `er_protocol_b()`
  use true held-out entity splits; `er_cv()` splits by entity.
- **Phase 3 (engineering)**: RNG hygiene; sparse-matrix guards (n > 5000);
  explicit export list; experimental neural code isolated to `R/experimental/`.
- **Phase 4 (classifiers)**: advisor's 11 supervised classifiers fused into
  the package (Firth-penalized logistic as default); validity gating with
  honest `valid = FALSE` fallbacks.

```{r session, eval = FALSE}
sessionInfo()
```
